A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-29 paper/writeup.tex
Finish The ARG -> an ARG updates
Closes #11

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees in which they
    do not represent coalescences.
    We show how these non-coalescing haplotypes (``unary regions'') may be
    inserted into the ancestral recombination graph (ARG), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting ARGs are smaller, faster to compute with,
    and the additional ancestral information that is inserted
    is nearly always correct where the initial ARG is correct.
    We provide efficient algorithms to infer unary regions within existing ARGs,
    and explore some consequences for ARGs inferred from real data.
    To do this, we introduce new metrics of agreement and disagreement between inferred ARGs
    that, unlike previous methods, consider the ARG as describing relationships between haplotypes
    rather than just a collection of trees.
\end{abstract}

@@

To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in the ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
Setting \texttt{coalescing\_segments\_only} to False includes this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

These statistics are also reflected at the genomic scale, using measures of dissimilarity.
Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
(Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
The top row shows the proportion of the given ARG that does not match the original (ARF),
showing that the total amount of mis-matching span
produced by extending haplotypes is very small ($\approx 1\%$).
(Simplification does not produce dissimilarity.)
The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
This shows that a large proportion of the ARG can be removed by simplification,
indicating that coalescent nodes are unary
over a substantial portion of their spans.
(Since the \texttt{simplify} operations removes these unary portions of haplotypes,
the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
the proportion of nodes' spans on which they are not unary.)
However, \algorithmref{alg:extend} can correctly replace most of these portions of haplotypes,
especially with larger sample sizes and sequence lengths
(the ``simplified--extended'' line; ``SE'').
For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
haplotypes are unary on about half their spans (on average),
and extending haplotypes can infer more than half of this span from coalescent information only.

>>>>

%%%%%%%%%%
\begin{abstract}
    Foreshadowing haplotype-based methods of the genomics era,
    it is an old observation that the ``junction'' between two distinct haplotypes
    produced by recombination is inherited as a Mendelian marker.
    In this paper, we describe how this recombination-mediated information
    can in many cases be recovered from inference based solely on polymorphic markers.
    In a genealogical context, this information reflects the persistence
    of ancestral haplotypes across marginal genealogical trees in which they
    do not represent coalescences.
    We show how these non-coalescing haplotypes (``unary regions'') may be
    inserted into ancestral recombination graphs (ARGs), a compact
    but information-rich data structure describing the genealogical
    relationships among recombinant sequences.
    The resulting ARGs are smaller, faster to compute with,
    and the additional ancestral information that is inserted
    is nearly always correct where the initial ARG is correct.
    We provide efficient algorithms to infer unary regions within existing ARGs,
    and explore some consequences for ARGs inferred from real data.
    To do this, we introduce new metrics of agreement and disagreement between ARGs
    that, unlike previous methods, consider ARGs as describing relationships between haplotypes
    rather than just a collection of trees.
\end{abstract}

@@

To do this, we simulate ARGs containing full haplotypes using \msprime{}
\citep{kelleher2016efficient,baumdicker2021efficient},
with the \texttt{coalescing\_segments\_only} option set to False.
Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
it only outputs information for nodes which contain a coalescence
(i.e., are the MRCA of some pair of samples at some point on the genome).
Furthermore, by default it only outputs those segments of the genome
on which there is a coalescence.
Said another way, by default all ancestral nodes in an ARG
output by \msprime{} are the MRCA of some pair of samples at every point in the genome
on which they are represented.
However, here we are interested in those segments of genome
on which the nodes are \emph{not} coalescent;
i.e., where they are unary in the marginal trees.
Setting \texttt{coalescing\_segments\_only} to False includes this information:
any ancestral segments for which these coalescent nodes
are ancestral to any samples -- so, the unary portions of their spans as well.
However, this includes more information than we want:
we hope to recover those portions of ancestral haplotypes
on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
coalesces with another spanning $[b, d)$, with $a < b < c < d$,
then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
to extend the node's span to $[a, b)$ and $[c, d)$
(on which the node is unary).
However, following this example, the first lineage might carry a segment $[x, y)$
that is disjoint from the segment $[a, d)$.
We call these segments ``isolated non-coalescent segments'';
they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
Such isolated segments will not be recovered by our algorithm,
and would likely be unrecoverable by any other method.
So, after simulation, we first remove these isolated, non-coalescent segments.
To give an idea of what proportion of the full spans of ancestral nodes
these isolated non-coalescent segments represent,
a simulation of 1000 samples
with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
has about half the total span of all nodes in isolated, non-coalescent segments.
For more discussion of these segments, see \citet{baumdicker2021efficient}.

@@

These statistics are also reflected at the genomic scale, using measures of dissimilarity.
Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
(Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
The top row shows the proportion of the given ARG that does not match the original (ARF),
showing that the total amount of mis-matching span
produced by extending haplotypes is very small ($\approx 1\%$).
(Simplification does not produce dissimilarity.)
The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
This shows that a large proportion of an ARG can be removed by simplification,
indicating that coalescent nodes are unary
over a substantial portion of their spans.
(Since the \texttt{simplify} operations removes these unary portions of haplotypes,
the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
the proportion of nodes' spans on which they are not unary.)
However, \algorithmref{alg:extend} can correctly replace most of these portions of haplotypes,
especially with larger sample sizes and sequence lengths
(the ``simplified--extended'' line; ``SE'').
For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
haplotypes are unary on about half their spans (on average),
and extending haplotypes can infer more than half of this span from coalescent information only.
